Discrete Quantum Theories and Computing

### 2019-01-26 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" and Acknowledgments Updating...

\usepackage{tabularx}
\newcolumntype{Y}{>{\centering\doublespacing\arraybackslash}X}

@@

I found the sentences in my Acknowledgments almost always contains
“thank”, and looks boring. To improve my Acknowledgments, I
briefly read the Acknowledgments in other theses \cite{Chih2013,Sun2015,Lee2013}.

@@

\todo{Confirm people to be acknowledged like the way being acknowledged\ldots{}
}I want to thank Prof.~Gerardo Ortiz, Prof.~Amr Sabry (\textarabic{عمرو صبري}),
and Prof.~Andrew Hanson for publishing three papers \cite{geometry2013,DQT2014,THOS2018}
with me, and enormous other things. These papers become the main part
of my thesis, especially most of Chapter~\ref{chap:DQT=000026DQC}
is based on ``Geometry of discrete quantum computing''~\cite{geometry2013}
and ``Discrete quantum theories''~\cite{DQT2014}; most of Chapter~\ref{chap:QIVPM}
is based on ``Quantum Interval-Valued Probability: Contextuality
and the Born Rule''~\cite{THOS2018}. Together with Prof.~Dylan
Thurston and Prof.~Shouhong Wang (汪守宏), I want to thank you for getting
together to understand what I did and passing my dissertation proposal
Spring 2017 and defense Fall 2018. Also thank Prof.\ Dylan Thurston
for serving in my advisory committee and my Tier 3 committee, with
other advisory committee members passing my Computer Science qualifying
exam Spring 2016. Especially, my survey of real computation in the
qualifying exam inspired me the comparison of quantum computer and
analog computer in Chapter~\ref{chap:Introduction}. Chapter~\ref{chap:Introduction}
also contains the motivation in my application for Rethinking Foundations
of Physics 2017 workshop reviewed by Prof.\ Amr Sabry~(\textarabic{عمرو صبري}).
I want to thank Prof.\ Lawrence Moss for inviting me to present our
results in the interdisciplinary logic seminar and theory seminar
in Computer Science, co-chairing my Tier 3 committee, writing an assessment
letter for me, and drawing my attention on Prof.\ Abramsky's paper~\cite{Abramsky2012},
which inspired me to merge quantum probability with discrete quantum
theories. And I thank John Gardiner for inspiring discussion on the
difficulties merging them~\cite{Gardiner2014}. These results are
improved and motivated the transition from DQT to QIVPM in Sec.~\ref{sec:Toward-IVPM}.
I also thank Prof.~Tom Lewis for helping me organize the unpublished
deterministic quantum algorithms in the term paper for SLST-T501 Academic
Writing course as presented in Sec.~\ref{subsec:Discrete-Deutsch-algorithm}.
I also thank Traci Nagle, Elizabeth ``Betsy'' Merceron, and Kexin
``Casey'' Chen (谌可心) for teaching me English in general so that
others can understand me more easily.

\todo{Type a table for acknowledgment so that I would have to write
``I thank'' so many times here? }Many people in both departments
of Mathematics and Computer Science need to do something more to make
my double-major works smoothly. I need to thank all of them, especially
directors of graduate study in Mathematics, especially Prof.\ Christopher
Judge and Prof.\ Matthias Weber; directors of Ph.D. study in Computer
Science, especially Prof.\ Yuqing Melanie Wu (吴愈青) and Prof.\ \textturkish{Funda
Ergun}.

Thank Kin Wai Chan for suggesting me the correct direction for Proposition~\ref{appH.sec},
Prof.\ 朱樺 for drawing my attention on Prof.\ Wan's book~\cite{Wan2006},
and Weihua Liu (刘伟华) for remaining me analysts call the right side
of the IVPM outer measure.

@@

\item My other AT\&T family plan members, especially Kexin ``Casey''
Chen (谌可心), Yao-yu Chih (池耀宇), Xinyi ``Diana'' Gong (龚欣怡), and Nanxi
``Nancy'' Zhang (张楠茜), for you always pay our telephone bill on
time so that\ldots\ldots\todo{ Maybe I don't have to thank them
as AT\&T family plan members, but I should thank some of them for
support and sharing information\ldots}

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick is made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
be an integer multiple of the elementary charge. A perfect square
cannot be drawn physically. Even we don't draw it, but merely ask
whether it exists in the space. This question is still beyond our
ability to answer because whether space itself is ultimately discrete
or continuous is still an unsolved question. Even if a perfect square
exists physically, we may not able to decide whether the ``square''
we discussed is really perfect or almost perfect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation closes to
zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and this difference is unlikely
to be compensated by error analysis techniques easily like the butterfly
effect. Therefore, it is is unlikely to utilize the realistic analog
computers to achieve the computational power predicted by this kind
of theoretical models.%
\begin{comment}
\todo{Maybe extended to consider probabilistic model? }
\end{comment}

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantity & Discrete & Discrete with extreme small units & Continuous, or no evidence to support they are discrete\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational device & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integer & Real number & Real number\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How does the theorical model predict the behavior of the physical device? & 
Reliably & 
Not reliably, but after we have better technology to manipulate the minimum units, we can more reliably predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be able compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum model &  & Quantum theories and computing over finite fields & Quantum interval-valued probability measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities for classical quantities inspires two types of quantum
models. Like its classical counterparts, quantum circuit model, the
most widely used model of quantum computing, manipulates quantities
which are not discrete apparently: the probability amplitudes of quantum
states, which are assumed in the field of complex numbers in the ``conventional
quantum theory'' (CQT)\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}. If the probability amplitudes are assumed to be ultimately discrete,
we choose to replace the field of complex numbers by discrete finite
fields for two reasons. On one hand, CQT is built upon the operations
among the probability amplitudes, and the probability amplitudes have
the required operations since the complex numbers form a field. On
the other hand, based on finite fields, we will define the fraction-like
cardinal probability which has extremely small units when the size
of the field is extremely large. Alternatively, even if the probability
amplitudes might never be found quantized, considering the precision
of quantities could still result in a better model. However, comparing
to the precision of probability amplitudes, it is easier to consider
the precision of their inducing probabilities because the idea of
``imprecise probabilities'' has well-studied classically \cite{Shafer1976,GilboaSchmeidler1994,Marinacci1999,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
\todo{Make sure whether realistic quantum computers could branch
their computation based on their probability amplitudes or measured
probabilities? If they can only branch on their measured probabilities,
it will provide another reason why we want to study the precision
of the inducing probabilities\ldots{} }This idea will be extended
to the quantum domain as quantum interval-valued probability measures
(QIVPM).

@@

\subsection{Classical and Quantum Probability Spaces}

>>>>

\usepackage{tabularx}
\renewcommand{\tabularxcolumn}[1]{m{#1}}
\newcolumntype{Y}{>{\centering\doublespacing\arraybackslash}X}
\newcolumntype{M}[1]{>{\centering\doublespacing\arraybackslash}m{\widthof{#1}}}

@@

I found the sentences in my Acknowledgments almost always contains
“thank”, and looks boring. To improve my Acknowledgments, I
briefly read the Acknowledgments in other theses \cite{Chih2013,Sun2015,Lee2013,Chen2016,sabry1995formal}.

@@

\begin{sidewaystable}
\noindent \begin{centering}
\caption{\label{tab:acknowledgments}I want to thank the professors co-authored
with me, suggested me as a committee member, and taught me, and the
collaboration or supervised works or studies become the basis of or
inspire this dissertation.}
\par\end{centering}
\noindent \centering{}\noindent \centering{}%
\begin{tabularx}{\linewidth}{M{MATH-M781}M{Co-author}M{(\textarabic{عمرو صبري})}M{Co-author}M{Thurston}M{Shouhong}M{Lawrence}Y}
\toprule 
\addlinespace
Committee or course & Gerardo Ortiz & Amr\nolinebreak[3] A. Sabry\linebreak[3] (\textarabic{عمرو صبري}) & Andrew\nolinebreak[3] J. Hanson & Dylan\nolinebreak[3] Paul Thurston & Shouhong Wang\linebreak[3] (汪守宏) & Lawrence\nolinebreak[3] S.\linebreak[2] Moss & Relation with the dissertation\tabularnewline\addlinespace
\addlinespace
\midrule
\midrule 
\addlinespace
 & Co-author & Co-author & Co-author &  &  &  & Most of Chapters~\ref{chap:Conventional-Quantum-Theory}, \ref{chap:Unrestricted-Finite-Fields},
and~\ref{chap:DQT=000026DQC} is based on \cite{geometry2013} and \cite{DQT2014}; most of Chapter~\ref{chap:QIVPM} is
based on \cite{THOS2018}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Research committee & Co-chair & Co-chair & Member & Co-chair & Member &  & Some questions in the dissertation proposal are answered in Chapter~\ref{chap:QIVPM},
and some of the remaining ones are explained in Chapter~\ref{chap:Further-Discussion}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
 &  & Suggested &  &  &  &  & The application for Rethinking Foundations of Physics 2017 workshop
inspires Chapter~\ref{chap:Introduction}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Advisory committee & Member & Member & Member & Member &  &  & The survey of real computation in the Computer Science qualifying
exam (chaired by Prof.\ Sabry) inspires Chapter~\ref{chap:Introduction}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
PHYS-P700 & Instructor &  &  &  &  &  & The textbook~\cite{Mermin2007} is
cited in Chapters~\ref{chap:Introduction}, \ref{chap:Conventional-Quantum-Theory},
\ref{chap:DQT=000026DQC}, and~\ref{chap:QIVPM}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
MATH-M781 &  &  &  &  &  & Instructor & The presented paper~\cite{Abramsky2012} is cited in Sec.~\ref{subsec:Classical-and-Quantum}.\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Tier 3 committee & Co-chair & Co-chair &  & Member &  & Co-chair & Two of
the assigned papers in the Tier 3 exam, \cite{Schumacher2012-SCHMQT}
and \cite{doi:10.1142/S0217984913500644},
are cited in Sec.~\ref{modalquantum}.\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
I want to thanks all my professors co-authored with me, suggested
me as a committee members, and taught me which become the bases or
inspired any part of my dissertation as listed in Table~\ref{tab:acknowledgments}.
Especially, Prof.\ Gerardo Ortiz, Prof.\ Amr A. Sabry\ (\textarabic{عمرو صبري}),
and Prof.\ Andrew J. Hanson meet me weekly for the pass few years,
read and tried to understand what I typed which sometimes even myself
cannot understand, taught me how to organize them in ``English''
so that general audience might be interested and have some chance
to understand, and encourage me to participate seminars, workshops,
and conferences. Prof.\ Lawrence Moss invited me to present our results
in the interdisciplinary logic seminar and theory seminar in Computer
Science and wrote an assessment letter for me. Although not directly
related to this dissertation, the knot theory studied with Prof.\ Dylan
Paul Thurston and the quantum interpretation discussed with Prof.\ Shouhong
Wang (汪守宏) also broadened my understanding on quantum theory and computing.
I would like express my appreciation to all of these.

Besides of Table~\ref{tab:acknowledgments}, I have some inspiring
discussion with John Gardiner about quantum probability measures over
finite fields~\cite{Gardiner2014}, and his results are improved
and motivated the transition from DQT to QIVPM in Sec.~\ref{sec:Toward-IVPM}.
Also, some unpublished computation for deterministic quantum algorithms
is organized in the term paper for SLST-T501 Academic Writing course.
I want to thank Prof.\ Tom Lewis for proof-reading them, and they
become the bases of Sec.~\ref{subsec:Discrete-Deutsch-algorithm}.
Besides of Prof.\ Tom Lewis, my other English teachers, especially
Traci Nagle, Elizabeth ``Betsy'' Merceron, and Kexin ``Casey''
Chen (谌可心), taught me a lot on expressing myself in English as well.
Others who help me indirectly on the content of dissertation including
Kin Wai Chan, Prof.\ 朱樺, Weihua Liu (刘伟华), 劉鈺平, Meng-Wei Chen
(陳孟瑋), \ldots , etc. This could be listed on and on, and I still
thank you even if your name is not listed here.\todo{Acknowledgments
cleaned up until here. }

Many people in both departments of Mathematics and Computer Science
need to do something more to make my double-major works smoothly.
I need to thank all of them, especially directors of graduate study
in Mathematics, especially Prof.\ Christopher Judge and Prof.\ Matthias
Weber; directors of Ph.D. study in Computer Science, especially Prof.\ Yuqing
Melanie Wu (吴愈青) and Prof.\ \textturkish{Funda Ergun}.

@@

\item My other AT\&T family plan members, especially Kexin ``Casey''
Chen (谌可心), Yao-yu Chih (池耀宇), and Xinyi ``Diana'' Gong (龚欣怡), for
you always pay our telephone bill on time so that\ldots\ldots\todo{
Maybe I don't have to thank them as AT\&T family plan members, but
I should thank some of them for support and sharing information\ldots}

@@

Although it is convenient to assume physical quantities is as continuous
as their mathematical models, each physical quantity discussed in
the previous paragraph could not have a simple one-to-one correspondence
with a real number. Chuang-tzu's stick is made by molecules and cannot
be cut in halves endlessly. The input charge of an integrator must
be an integer multiple of the elementary charge. A perfect square
cannot be drawn physically. Even we don't draw it, but merely ask
whether it exists in the space. This question is still beyond our
ability to answer because whether space itself is ultimately discrete
or continuous is still an unsolved question. Even if a perfect square
exists physically, we may not able to decide whether the ``square''
we discussed is really perfect or almost perfect differing only by
an arbitrarily small extent. Similar situation applying to the length
of a slide rule, it is beyond our ability to decide whether its length
is ultimately discrete or continuous. Even if its length is continuous,
identifying which real number represents its length requires infinite
precision, but the precision in reality is limited generally three
or four significant figures~\cite{wiki:AnalogComputer}. As the results,
even if analog computers really store real numbers, they cannot be
precisely read, written, and used to branch the computation. The opposite
of the last statement is sometimes assumed by real computational models.
For example, Blum, Cucker, Shub, and Smale modeled their BCSS machine
to allow the operations of deciding whether a number is greater or
equal to zero over real numbers and whether a number is exactly zero
over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Hence, when the quantities used to branch the computation closes to
zero, the branch chosen by a BCSS machine cannot reliably predict
the branch chosen by a realistic analog computer. Since two branches
of computations may not have any relations, the follow-up computational
paths may be significantly different, and this difference is unlikely
to be compensated by error analysis techniques easily like the butterfly
effect. Therefore, we might not be able to utilize a realistic analog
computer to acquire the computational power predicted by this kind
of theoretical models.%
\begin{comment}
\todo{Maybe extended to consider probabilistic model? }
\end{comment}

@@

\centering{}\centering{}%
\begin{tabularx}{\linewidth}{YcYY}
\toprule 
\addlinespace
Physical quantity & Discrete & Discrete with extreme small units & Continuous, or no evidence to support they are discrete\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational device & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Mathematical representation & Integer & Real number & Real number\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How does the theorical model predict the behavior of the physical device? & 
Reliably & 
Not reliably, but after we have better technology to manipulate the minimum units, we can more reliably predict the physical phenomenon and their computational power by a better discrete model. & 
Not reliably, because precision can never be high enough, and the difference between their computational power might not be able compensated by error analysis techniques. \tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Inspired quantum model &  & Quantum theories and computing over finite fields & Quantum interval-valued probability measures (QIVPM)\tabularnewline\addlinespace
\addlinespace
\bottomrule
\end{tabularx}
\end{sidewaystable}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and two
possibilities for classical quantities inspires two types of quantum
models. Like its classical counterparts, quantum circuit model, the
most widely used model of quantum computing, manipulates quantities
which are not discrete apparently: the probability amplitudes of quantum
states, which are assumed in the field of complex numbers in the ``conventional
quantum theory'' (CQT)\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}. If the probability amplitudes are assumed to be ultimately discrete,
we choose to replace the field of complex numbers by discrete finite
fields for two reasons. On one hand, CQT is built upon the operations
among the probability amplitudes, and the probability amplitudes have
the required operations since the complex numbers form a field. On
the other hand, based on finite fields, we will define the fraction-like
cardinal probability which has extremely small units when the size
of the field is extremely large. Alternatively, even if the probability
amplitudes might never be found quantized, considering the precision
of quantities could still result in a better model. However, comparing
to the precision of probability amplitudes, it is easier to consider
the precision of their inducing probabilities because the idea of
``imprecise probability'' has well-studied classically \cite{Shafer1976,GilboaSchmeidler1994,Marinacci1999,JamisonLodwick2004,HuberRonchetti2009,Grabisch2016}.
\begin{comment}
\todo{Make sure whether realistic quantum computers could branch
their computation based on their probability amplitudes or measured
probabilities? If they can only branch on their measured probabilities,
it will provide another reason why we want to study the precision
of the inducing probabilities\ldots{} }
\end{comment}
This idea will be extended to quantum domain as quantum interval-valued
probability measure (QIVPM).

@@

\subsection{Classical and Quantum Probability Spaces\label{subsec:Classical-and-Quantum}}
